---
title: "Leslie-Caitlin-ADA-HOMEWORK-03"
author: "Caitlin Leslie"
date: "March 20, 2019"
output: html_document
---

#Problem 1#

Write a simple R function you call Z.prop.test(), that can perform one- or two-sample Z-tests for proportion data, using the following guidelines.

-Your function should take the following arguments: p1 and n1 (no default) representing the estimated proportion and sample size (e.g., based on your sample data); p2 and n2 (both defaulting to NULL) that contain a second sample’s proportion and sample size data in the event of a two-sample test; p0 (no default) as the expected value for the population proportion; and alternative (default “two.sided”) and conf.level (default 0.95), to be used in the same way as in the function t.test().

-When conducting a two-sample test, it should be p1 that is tested as being smaller or larger than p2 when alternative=“less” or alternative=“greater”, respectively, the same as in the use of x and y in the function t.test().

-The function should perform a one-sample Z-test using p1, n1, and p0 if either p2 or n2 (or both) is NULL.

-The function should contain a check for the rules of thumb we have talked about (n×π>5 and n×(1−π)>5) to ensure the validity of assuming the normal distribution in both the one- and two-sample settings. If this is violated, the function should still complete, but it should also print an appropriate warning message.

-The function should return a list containing the following elements: Z (the test statistic), P (the appropriate p value), and CI (the two-sided CI with respect to “conf.level” around p1 in the case of a one-sample test and around p2-p1 in the case of a two-sample test). For all test alternatives (“two.sided”, “greater”, “less”), calculate symmetric CIs based on quantiles of the normal distribution rather than worrying about calculating single-limit confidence bounds.


```{r}

Z.prop.test<-function(p1, n1, p2=NULL, n2=NULL, p0, alternative="two.sided", conf.level=0.95) 
{ if(n1*p0<5){Warning("Normal Approximation Not Appropriate")}
  if(n1*(1-p0)<5) {Warning("Normal Approximation Not Appropriate")}
  if(is.null(p2)){
    Z<-(p1-p0)/sqrt(p0*(1-p0)/n1) #One-sample Z-test
    Z
    
    lower <- p1 - qnorm(0.975) * sqrt(p1 * (1 - p1)/n1)
    upper <- p1 + qnorm(0.975) * sqrt(p1 * (1 - p1)/n1)
    ci <- c(lower, upper)
    ci}
  
    
  else { #Assuming p2 has a value
    pi=0
    pstar=((p1*n1)+(p2*n2))/(n1+n2)
    Z<-(p2-p1-pi)/sqrt((pstar*(1-pstar)*((1/n1)+(1/n2)))) #Two-sample Z-test
    Z
    
    lower <- (p2-p1) - qnorm(0.975) * sqrt(((p2*(1-p2))/n2) + ((p1*(1-p1))/n1))
    upper <- (p2-p1) + qnorm(0.975) * sqrt(((p2*(1-p2))/n2) + ((p1*(1-p1))/n1))
    ci <- c(lower, upper)
    ci
  }
    
  P.upper<-1-pnorm(Z, lower.tail=TRUE)
  P.lower<-pnorm(Z, lower.tail=FALSE)
  P <- P.upper + P.lower
  P
  
  if(alternative=="greater") {
  P <- pnorm(Z, lower.tail=FALSE)
  P
  }

  if(alternative=="less") {
    P <- pnorm(z, lower.tail = TRUE)
    P  
  }
}
#I keep getting NULL when I try to run this on one of our examples in class, and I don't know why, but this is the best I could come up with
```


#Problem 2#

The comparative primate dataset we have used from Kamilar and Cooper has in it a large number of variables related to life history and body size. For this exercise, the end aim is to fit a simple linear regression model to predict longevity (MaxLongevity_m) measured in months from species’ brain size (Brain_Size_Species_Mean) measured in grams. Do the following for both longevity~brain size and log(longevity)~log(brain size).

```{r}
library(curl)
library(ggplot2)
f <- curl("https://raw.githubusercontent.com/difiore/ADA-2019/master/KamilarAndCooperData.csv")
d <- read.csv(f, header = TRUE, stringsAsFactors = FALSE)

d<-d[complete.cases(d[ , c(5,20)]),]  #removed all individuals who had an NA for either MaxLongevity_m or Brain_Size_Species_Mean ( I think - I found this code on StackOverflow, but it seems to work)
```



***For Longevity~Brain size:***

* Fit the regression model and, using {ggplot2}, produce a scatterplot with the fitted line superimposed upon the data. Append the the fitted model equation to your plot (HINT: use the function geom_text()).

Finding the covariance

```{r}

l<-d$MaxLongevity_m
b<-d$Brain_Size_Species_Mean

n<-length(l)

cov_bl<-sum(((b-mean(b))*(l-mean(l)))/(n-1))
cov_bl

```

Finding the correlation coefficient

```{r}
sd_b<-sd(b)
sd_l<-sd(l)

cor_bl <- cov_bl/(sd_b * sd_l)
cor_bl

```

Shifting the mean from each value to "center" each variable

```{r}
y <- l - mean(l)
x <- b - mean(b)
z<-data.frame(cbind(x,y))
g <- ggplot(data = z, aes(x = x, y = y)) + geom_point()
g

```

Finding B1

```{r}
beta1<-cor_bl*(sd_l/sd_b)
beta1
```

Finding B0

```{r}

beta0<-mean(l)-beta1*mean(b)
beta0

```

Alternatively, we can use the lm() function

```{r}

mlb<-lm(l~b, data=d)
mlb
```

Plot with fitted line superimposed

```{r}

g <- ggplot(data = d, aes(x = b, y = l))
g <- g + geom_point()
g <- g + geom_smooth(method = "lm", formula = y ~ x)
g <- g + labs(x="Brain Size", y = "Longevity")
g <- g + geom_text(x=225, y= 700, label= "y = 1.218x + 248.952")
g

```



* Identify and interpret the point estimate of the slope (β1), as well as the outcome of the test associated with the hypotheses H0: β1=0; HA: β1≠0. Also, find a 90% CI for the slope (β1) parameter.

```{r}
beta1<-cor_bl*(sd_l/sd_b)
beta1
```


This point is the slope of the Ordinary Least Squares regression line for longevity as a function of brain size. It predicts that for every 1 gram increase in brain size, there will be a 1.21799 months increase in longevity. 

If the null was true, and B1=0, there would be no relationship between brain size and longevity, and you would not be able to predict longevity from brain size. If the alterative was true, and B1 =/= 0, then there would be a relationship, either positive or negative, between the two variables, and you would be able to predict longevity from brain size.


```{r}
summary(mlb)

```
From the summary of the linear model, we can see that the p=value of B1 is less than 2e-16, which supports the alternative hypothesis that B1 does not equal 0.

```{r}
alpha<-0.10
mlbCI<-confint(mlb,level=1-alpha)
mlbCI
```
So, the 90% confidence interval for the slope parameter is 1.035571 to 1.40041.

* Using your model, add lines for the 90% confidence and prediction interval bands on the plot, and add a legend to differentiate between the lines.

```{r}
#Make a data table that includes predicted values of y
l_hat<-predict(mlb, newdata=data.frame(brain_size=b))
df<-data.frame(cbind(b,l,l_hat))
names(df)<-c("x", "y", "yhat")
head(df)


#Calculate 90% confidence interval 
ci <- predict(mlb, newdata = data.frame(brain_size = b), interval = "confidence", 
    level = 0.90)  

#Calculate prediction interval
pi <- predict(mlb, newdata = data.frame(brain_size = b), interval = "prediction", 
    level = 0.90)  


#Bind into one data frame
df<-cbind(df,ci,pi)
names(df)<-c("x","y","yhat","CIfit", "CIlwr", "CIupr", "PIfit", "PIlwr", "PIupr")
head(df)


#Plot

g <- ggplot(data = df, aes(x = x, y = y))
g <- g + geom_point(alpha=0.10)
g <- g + geom_line(aes(x = x, y = CIfit), colour = "black")
g <- g + geom_line(aes(x = x, y = CIlwr, color="Confidence Interval"))
g <- g + geom_line(aes(x = x, y = CIupr, color="Confidence Interval"))
g <- g + geom_line(aes(x = x, y = PIlwr, color="Prediction Interval"))
g <- g + geom_line(aes(x = x, y = PIupr, color="Prediction Interval"))
g <- g + scale_color_manual(values = c("blue", "red"))
g

```


* Produce a point estimate and associated 90% prediction interval for the longevity of a species whose brain weight is 800 gm. Do you trust the model to predict observations accurately for this value of the explanatory variable? Why or why not?

```{r}

ci_l <- predict(mlb, newdata = data.frame(b = 800), interval = "prediction", 
    level = 0.90)  
ci_l

```

So the point estimate is 1223.345 and the 90% prediction interval is 1021.805 to 1424.884.  I would not trust this model to accurately predict observations for this value because the 90% confidence interval ranges over 400 months, which is a significant variation.



***For log(longevity) ~ log(brain size)***

* Fit the regression model and, using {ggplot2}, produce a scatterplot with the fitted line superimposed upon the data. Append the the fitted model equation to your plot (HINT: use the function geom_text()).


Finding the covariance

```{r}

l<-log(d$MaxLongevity_m)
b<-log(d$Brain_Size_Species_Mean)

n<-length(l)

cov_bl<-sum(((b-mean(b))*(l-mean(l)))/(n-1))
cov_bl

```

Finding the correlation coefficient

```{r}
sd_b<-sd(b)
sd_l<-sd(l)

cor_bl <- cov_bl/(sd_b * sd_l)
cor_bl

```

Shifting the mean from each value to "center" each variable

```{r}
y <- l - mean(l)
x <- b - mean(b)
z<-data.frame(cbind(x,y))
g <- ggplot(data = z, aes(x = x, y = y)) + geom_point()
g

```

Finding B1

```{r}
beta1<-cor_bl*(sd_l/sd_b)
beta1
```

Finding B0

```{r}

beta0<-mean(l)-beta1*mean(b)
beta0

```

Alternatively, we can use the lm() function

```{r}

mlb<-lm(l~b, data=d)
mlb
```

Plot with fitted line superimposed

```{r}

g <- ggplot(data = d, aes(x = b, y = l))
g <- g + geom_point()
g <- g + geom_smooth(method = "lm", formula = y ~ x)
g <- g + labs(x="log(Brain Size)", y="log(Longevity)")
g <- g + geom_text(x=3, y= 6.25, label= "y = 0.234x + 4.879")
g

```

* Identify and interpret the point estimate of the slope (β1), as well as the outcome of the test associated with the hypotheses H0: β1=0; HA: β1≠0. Also, find a 90% CI for the slope (β1) parameter.


```{r}
beta1<-cor_bl*(sd_l/sd_b)
beta1
```


This point is the slope of the Ordinary Least Squares regression line for longevity as a function of brain size. It predicts that for every 1 gram increase in brain size, there will be about a 0.234 months increase in longevity. 

If the null was true, and B1=0, there would be no relationship between brain size and longevity, and you would not be able to predict longevity from brain size. If the alterative was true, and B1 =/= 0, then there would be a relationship, either positive or negative, between the two variables, and you would be able to predict longevity from brain size.


```{r}
summary(mlb)

```
From the summary of the linear model, we can see that the p=value of B1 is less than 2e-16, which supports the alternative hypothesis that B1 does not equal 0.

```{r}
alpha<-0.10
mlbCI<-confint(mlb,level=1-alpha)
mlbCI
```
So, the 90% confidence interval for the slope parameter is 0.2046396 to 0.2636595.


* Using your model, add lines for the 90% confidence and prediction interval bands on the plot, and add a legend to differentiate between the lines.

```{r}
#Make a data table that includes predicted values of y
l_hat<-predict(mlb, newdata=data.frame(log_brain_size=b))
df<-data.frame(cbind(b,l,l_hat))
names(df)<-c("x", "y", "yhat")
head(df)


#Calculate 90% confidence interval 
ci <- predict(mlb, newdata = data.frame(brain_size = b), interval = "confidence", 
    level = 0.90)  

#Calculate prediction interval
pi <- predict(mlb, newdata = data.frame(brain_size = b), interval = "prediction", 
    level = 0.90)  


#Bind into one data frame
df<-cbind(df,ci,pi)
names(df)<-c("x","y","yhat","CIfit", "CIlwr", "CIupr", "PIfit", "PIlwr", "PIupr")
head(df)


#Plot

g <- ggplot(data = df, aes(x = x, y = y))
g <- g + geom_point(alpha=0.10)
g <- g + geom_line(aes(x = x, y = CIfit), colour = "black")
g <- g + geom_line(aes(x = x, y = CIlwr, color="Confidence Interval"))
g <- g + geom_line(aes(x = x, y = CIupr, color="Confidence Interval"))
g <- g + geom_line(aes(x = x, y = PIlwr, color="Prediction Interval"))
g <- g + geom_line(aes(x = x, y = PIupr, color="Prediction Interval"))
g <- g + scale_color_manual(values = c("blue", "red"))
g

```


* Produce a point estimate and associated 90% prediction interval for the longevity of a species whose brain weight is 800 gm. Do you trust the model to predict observations accurately for this value of the explanatory variable? Why or why not?

```{r}

ci_l <- predict(mlb, newdata = data.frame(b = 800), interval = "prediction", 
    level = 0.90)  
ci_l

```
So the point estimate is 192.1986 and the 90% prediction interval is 168.6955 to 215.7017.  I would somewhat trust this model to accurately predict observations for this value because the 90% confidence interval ranges over about 40 months, which is still a pretty wide range, but is much shorter than the previous (non logarithmic) model.

* Looking at your two models, which do you think is better? Why?

The logarithmic model seems better, because it gives a much narrower confidence interval. 


